---
title: "Getting Started with rollama"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{getting-started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The `rollama` package provides text classification with constrained output and confidence scoring using multiple inference backends: **Ollama**, **vLLM**, **SGLang**, and **llama.cpp**.

## Installation

```r
# Install from GitHub
# install.packages("pak")
pak::pak("paluigi/rollama-classifier")
```

### Prerequisites

**Ollama backend**: Install [Ollama](https://ollama.com/download) and pull a model:

```bash
ollama pull llama3.2
```

Recommended models: `llama3.2`, `llama3.1`, `mistral`, `qwen2.5`.

## Quick Start

### Ollama Backend (Original)

```{r eval=FALSE}
library(rollama)

classifier <- ollama_classifier("llama3.2")

result <- classify(
  classifier,
  text = "I love this product!",
  choices = c("positive", "negative", "neutral")
)

print(result$prediction)
#> "positive"
print(result$confidence)
#> 0.85
```

### vLLM Backend

```{r eval=FALSE}
backend <- vllm_backend("meta-llama/Llama-3.2-3B-Instruct")
classifier <- llm_classifier(backend)

result <- classify(
  classifier,
  text = "I love this product!",
  choices = c("positive", "negative", "neutral")
)
```

## Usage Patterns

### Basic Classification

Classify text into predefined choices:

```{r eval=FALSE}
classifier <- ollama_classifier("llama3.2")

result <- classify(
  classifier,
  text = "The goalkeeper made an incredible save!",
  choices = c("sports", "politics", "technology", "entertainment")
)
```

### Classification with Label Descriptions

Providing descriptions helps the model understand each category better:

```{r eval=FALSE}
choices <- list(
  positive = "Text expresses happiness, satisfaction, or approval",
  negative = "Text expresses anger, disappointment, or disapproval",
  mixed = "Text contains both positive and negative sentiments",
  neutral = "Text is factual without strong emotional content"
)

result <- classify(
  classifier,
  text = "The food was amazing but the service was terrible.",
  choices = choices
)
```

### Custom System Prompt

Override the default system prompt for specialized tasks:

```{r eval=FALSE}
result <- classify(
  classifier,
  text = "The quarterly earnings exceeded analyst expectations.",
  choices = c("bullish", "bearish", "neutral"),
  system_prompt = "You are a financial sentiment analyzer.
                   Classify financial news based on market sentiment."
)
```

### Generate Only (Fastest)

When you only need the prediction without confidence scores:

```{r eval=FALSE}
prediction <- generate(
  classifier,
  text = "The team won the championship!",
  choices = c("sports", "finance", "politics")
)
```

### Batch Classification

Classify multiple texts efficiently:

```{r eval=FALSE}
texts <- c(
  "The goalkeeper made an incredible save!",
  "The central bank raised interest rates.",
  "The new smartphone features a revolutionary camera."
)

results <- batch_classify(
  classifier,
  texts = texts,
  choices = c("sports", "finance", "technology")
)

purrr::map2_chr(texts, results, ~ paste0(.x, " -> ", .y$prediction))
```

### Custom System Prompt with Batch

```{r eval=FALSE}
results <- batch_classify(
  classifier,
  texts = texts,
  choices = c("bullish", "bearish", "neutral"),
  system_prompt = "You are a financial sentiment analyzer."
)
```

## Choosing a Method

| Use Case                              | Recommended Method                    |
|---------------------------------------|----------------------------------------|
| Speed is critical, no confidence       | `generate()`                           |
| Confidence scores needed               | `classify()`                           |
| Batch processing                       | `batch_classify()` or `batch_generate()`  |

## Sample Data

The package ships with a sample dataset of 20 customer support tickets:

```{r}
data(sample_tickets, package = "rollama")
head(sample_tickets)
```

```{r eval=FALSE}
# Test classification accuracy
predictions <- batch_generate(
  classifier,
  texts = sample_tickets$text,
  choices = c("billing", "technical_support", "account", "general")
)

mean(predictions == sample_tickets$expected_label)
```
