---
title: "Getting Started with rollama"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{getting-started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The `rollama` package provides text classification with constrained
output and calibrated confidence scoring using four inference backends:
**Ollama**, **vLLM**, **SGLang**, and **llama.cpp**.

Every backend is constructed with a backend function and wrapped in a
single `llm_classifier()`. The classifier then offers two scoring
methods — `generate()` and `classify()` — both returning a
`classification_result` with a prediction, confidence, and a full
probability distribution.

## Installation

```r
# Install from GitHub
# install.packages("pak")
pak::pak("paluigi/rollama-classifier")
```

### Prerequisites

**Ollama backend**: Install [Ollama](https://ollama.com/download) (>=v0.12)
and pull a model:

```bash
ollama pull llama3.2
```

Recommended models: `llama3.2`, `llama3.1`, `mistral`, `qwen2.5`.

## Quick Start

### Ollama Backend

```{r eval=FALSE}
library(rollama)

backend <- ollama_backend(model = "llama3.2")
classifier <- llm_classifier(backend)

result <- classify(
  classifier,
  text = "I love this product!",
  choices = c("positive", "negative", "neutral")
)

print(result$prediction)
#> "positive"
print(result$confidence)
#> 0.85
```

### vLLM Backend

```{r eval=FALSE}
backend <- vllm_backend("meta-llama/Llama-3.2-3B-Instruct")
classifier <- llm_classifier(backend)

result <- classify(
  classifier,
  text = "I love this product!",
  choices = c("positive", "negative", "neutral")
)
```

## `generate()` vs `classify()`

rollama offers two complementary scoring methods. Both return a
`classification_result`; they differ in how they spend API calls and how
their confidence is computed.

| | `generate()` | `classify()` |
|---|---|---|
| **Strategy** | Hierarchical constrained generation with reproportioning | Multi-call completion scoring |
| **API calls** | 1 to `max_calls` (budget-controlled) | N calls for N labels |
| **Confidence** | Divergence-aware; exact when fully resolved | Always exact (calibrated) |
| **Normalization** | Geometric mean of per-token logprobs | Geometric mean of per-token logprobs |
| **`method`** | `"adaptive_generate"` | `"multi_call"` |
| **`approximate`** | `TRUE` if any label has partial coverage | Always `FALSE` |

- **`generate()`** makes one or a few constrained calls. The first call
  constrains the model to *all* labels and produces an internally
  consistent probability distribution from divergence-aware logprobs along
  the winning generation path. Supplementary calls (when `max_calls > 1`)
  resolve clusters of labels that share a token prefix by **reproportioning**
  probability mass *within* each cluster — they never change between-group
  totals, so accuracy never degrades as the call budget grows. The result
  may be *approximate* when the budget runs out before every label is fully
  resolved.
- **`classify()`** scores every label independently via the backend's
  completion-scoring endpoint, then applies a softmax over
  geometric-mean-normalized scores. It always makes N calls for N labels
  and is always exact.

### `max_calls`

The `max_calls` argument controls `generate()`'s API budget. The first
call is always made over *all* labels; supplementary calls only
reproportion probability mass within unresolved clusters:

| `max_calls` | Behavior | `approximate` |
|---|---|---|
| `1` (default) | Single constrained call; no cluster resolution | Possibly `TRUE` |
| `K` (integer) | Adaptive: resolves label clusters up to K calls | Possibly `TRUE` |
| `NULL` | Resolves all clusters recursively; exact | `FALSE` |

```{r eval=FALSE}
# Fast, single-call prediction (may be approximate)
result <- generate(classifier, text = txt, choices = choices, max_calls = 1)
print(result$approximate) # TRUE if any label was only partially scored

# Adaptive: spend up to 3 calls to resolve ambiguities
result <- generate(classifier, text = txt, choices = choices, max_calls = 3)

# Exact: resolve every label fully
result <- generate(classifier, text = txt, choices = choices, max_calls = NULL)
```

### `coverage` and `approximate`

When `generate()` cannot fully resolve a label within the call budget,
that label's token logprobs are only partially known. Two fields report
this:

- **`result$coverage`** — a named numeric vector giving the fraction of
  tokens scored per label (0.0 to 1.0). `1.0` means the label is fully
  resolved.
- **`result$approximate`** — `TRUE` if *any* label has coverage below
  1.0. For `classify()` this is always `FALSE`.

```{r eval=FALSE}
result <- generate(classifier, text = txt, choices = choices, max_calls = 1)

print(result$coverage)
#>   positive    negative     neutral
#>         1.0         0.5         1.0
print(result$approximate)
#> TRUE
print(result$n_calls)
#> 1
```

## Usage Patterns

### Basic Classification

Classify text into predefined choices:

```{r eval=FALSE}
backend <- ollama_backend(model = "llama3.2")
classifier <- llm_classifier(backend)

result <- classify(
  classifier,
  text = "The goalkeeper made an incredible save!",
  choices = c("sports", "politics", "technology", "entertainment")
)
```

### Classification with Label Descriptions

Providing descriptions helps the model understand each category better:

```{r eval=FALSE}
choices <- list(
  positive = "Text expresses happiness, satisfaction, or approval",
  negative = "Text expresses anger, disappointment, or disapproval",
  mixed = "Text contains both positive and negative sentiments",
  neutral = "Text is factual without strong emotional content"
)

result <- classify(
  classifier,
  text = "The food was amazing but the service was terrible.",
  choices = choices
)
```

### Custom System Prompt

Override the default system prompt for specialized tasks:

```{r eval=FALSE}
result <- classify(
  classifier,
  text = "The quarterly earnings exceeded analyst expectations.",
  choices = c("bullish", "bearish", "neutral"),
  system_prompt = "You are a financial sentiment analyzer.
                   Classify financial news based on market sentiment."
)
```

### Hierarchical Generation

`generate()` returns a `classification_result` just like `classify()`,
reconstructed from constrained generation:

```{r eval=FALSE}
result <- generate(
  classifier,
  text = "The team won the championship!",
  choices = c("sports", "finance", "politics"),
  max_calls = 1
)

print(result$prediction)
print(result$confidence)
print(result$method)      # "adaptive_generate"
print(result$approximate) # TRUE if any label is partially scored
```

### Batch Classification

Classify multiple texts efficiently. The batch variants return a list of
`classification_result` objects:

```{r eval=FALSE}
texts <- c(
  "The goalkeeper made an incredible save!",
  "The central bank raised interest rates.",
  "The new smartphone features a revolutionary camera."
)

results <- batch_classify(
  classifier,
  texts = texts,
  choices = c("sports", "finance", "technology")
)

purrr::map2_chr(texts, results, ~ paste0(.x, " -> ", .y$prediction))
```

### Custom System Prompt with Batch

```{r eval=FALSE}
results <- batch_classify(
  classifier,
  texts = texts,
  choices = c("bullish", "bearish", "neutral"),
  system_prompt = "You are a financial sentiment analyzer."
)
```

## The `classification_result` Object

Both methods return the same structure:

```{r eval=FALSE}
result$prediction      # Character: predicted label
result$confidence      # Numeric (0-1): confidence score
result$probabilities   # Named numeric: probability distribution (sums to 1)
result$method          # "adaptive_generate" or "multi_call"
result$approximate     # Logical: TRUE if any label has partial coverage
result$coverage        # Named numeric: per-label token coverage (0.0-1.0)
result$n_calls         # Integer: number of API calls made
result$raw_response    # List: raw data for debugging
```

## Choosing a Method

| Use Case | Recommended Method |
|---------------------------------------|----------------------------------------|
| Speed is critical, limited call budget | `generate(max_calls = 1)` |
| Balanced speed and accuracy | `generate(max_calls = K)` |
| Exact confidence on a budget | `generate(max_calls = NULL)` |
| Maximum calibration accuracy | `classify()` |
| Batch processing | `batch_classify()` or `batch_generate()` |

## Sample Data

The package ships with a sample dataset of 20 customer support tickets:

```{r}
data(sample_tickets, package = "rollama")
head(sample_tickets)
```

```{r eval=FALSE}
# Test classification accuracy
results <- batch_classify(
  classifier,
  texts = sample_tickets$text,
  choices = c("billing", "technical_support", "account", "general")
)

predictions <- purrr::map_chr(results, "prediction")
mean(predictions == sample_tickets$expected_label)
```
