---
title: "Inference Backends"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{inference-backends}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Overview

`rollama` supports four inference backends, all accessed through the
same `llm_classifier()` interface. Each backend is created by a
constructor function and exposes a unified set of methods: `chat()`,
`score()`, and `tokenize()`.

```r
library(rollama)

backend   <- ollama_backend("llama3.2")   # or vllm / sglang / llamacpp
classifier <- llm_classifier(backend)
```

Because the classifier talks to backends through this common interface,
`generate()`, `classify()`, and their batch variants work identically no
matter which engine you use. The backends differ only in how they run
and how they constrain output to valid labels.

## Backend capabilities and label constraints

Every backend constrains the model's generation to one of the supplied
labels so that `generate()` always produces a valid choice. The
mechanism differs per engine, captured by the
`supports_bare_label_constraint` flag:

| Backend | `supports_bare_label_constraint` | Constraint mechanism | Output form |
|---|---|---|---|
| `ollama_backend()`  | `FALSE` | JSON Schema `enum` (via `format`) | `{"label": "..."}` |
| `vllm_backend()`    | `TRUE`  | `guided_choice`                   | bare label text |
| `sglang_backend()`  | `TRUE`  | `regex`                           | bare label text |
| `llamacpp_backend()`| `TRUE`  | GBNF `grammar`                    | bare label text |

Backends with `supports_bare_label_constraint = TRUE` emit the label as
plain text. The Ollama backend wraps the label in a small JSON object
(`{"label": "<chosen>"}`); `rollama` filters the structural JSON tokens
during trie reconstruction and uses context-dependent tokenization so
the prefix trie matches the actual response tokens. This is an internal
detail — both paths feed the same `generate()` / `classify()` logic.

## Ollama

The default backend. Uses the native Ollama REST API (`/api/chat`,
`/api/generate`, `/api/tokenize`). Requires the Ollama runtime (>=v0.12)
installed locally.

**Start:**

```bash
ollama pull llama3.2
ollama serve
```

**Connect:**

```{r eval=FALSE}
backend <- ollama_backend(model = "llama3.2")
classifier <- llm_classifier(backend)
```

`ollama_backend()` defaults to `host = "http://localhost:11434"`. To
point at a remote Ollama instance, set `host`:

```{r eval=FALSE}
backend <- ollama_backend(
  model = "llama3.2",
  host  = "https://my-ollama.example.com"
)
```

Ollama constrains labels with a JSON Schema `enum` passed via the
`format` parameter, so it sets `supports_bare_label_constraint = FALSE`.

## vLLM

High-throughput serving engine for LLMs. Communicates via the
OpenAI-compatible API and supports guided decoding and logprobs out of
the box.

**Start a local server:**

```bash
python -m vllm.entrypoints.openai.api_server \
    --model meta-llama/Llama-3.2-3B-Instruct \
    --host 0.0.0.0 --port 8000
```

**Connect:**

```{r eval=FALSE}
backend <- vllm_backend(
  model = "meta-llama/Llama-3.2-3B-Instruct",
  base_url = "http://localhost:8000/v1"
)
classifier <- llm_classifier(backend)
```

vLLM constrains labels natively with `guided_choice`, producing bare
label text, so `supports_bare_label_constraint = TRUE`.

**Remote server:**

```{r eval=FALSE}
backend <- vllm_backend(
  model = "your-model",
  base_url = "https://your-vllm-server.com/v1",
  api_key = "your-api-key"
)
classifier <- llm_classifier(backend)
```

## SGLang

Fast serving system for large language models with efficient radix
attention. Also OpenAI-compatible.

**Start a local server:**

```bash
python -m sglang.launch_server \
    --model-path meta-llama/Llama-3.2-3B-Instruct \
    --host 0.0.0.0 --port 30000
```

**Connect:**

```{r eval=FALSE}
backend <- sglang_backend(
  model = "meta-llama/Llama-3.2-3B-Instruct",
  base_url = "http://localhost:30000/v1"
)
classifier <- llm_classifier(backend)
```

SGLang constrains labels with a `regex` built from the label set
(`(label1|label2|...)`), producing bare label text, so
`supports_bare_label_constraint = TRUE`.

## llama.cpp

Lightweight inference via `llama-server`. Ideal for CPU or mixed
CPU/GPU environments.

**Start a local server:**

```bash
./llama-server -m model.gguf --host 0.0.0.0 --port 8080 -c 4096
```

**Connect:**

```{r eval=FALSE}
backend <- llamacpp_backend(
  model = "model",
  base_url = "http://localhost:8080/v1"
)
classifier <- llm_classifier(backend)
```

llama.cpp constrains labels with a GBNF `grammar`
(`root ::= "label1" | "label2" | "label3"`), producing bare label text,
so `supports_bare_label_constraint = TRUE`.

> **Note:** Logprobs support requires `llama-server` to be compiled with
> the appropriate flag (e.g. `LLAMA_SUPPORT_LOGPROBS`). Constrained
> generation via grammar is available in recent llama.cpp builds.

## Backend Configuration

All backends share common configuration options:

| Parameter    | Default          | Description                                |
|--------------|------------------|--------------------------------------------|
| `model`      | *(required)*     | Model identifier                           |
| `base_url` / `host` | Engine-specific | Base URL of the inference server    |
| `api_key`    | `"not-needed"`   | API key for authentication                 |
| `timeout`    | `120`            | Request timeout in seconds                 |
| `max_tokens` | `256`            | Maximum tokens to generate                 |
| `extra_body` | `{}`             | Extra parameters merged into every request |

> Ollama uses `host` instead of `base_url` and has no `api_key`
> parameter (the native API needs none). The other backends use
> `base_url` and `api_key` to match the OpenAI-compatible convention.

## Switching Backends

`llm_classifier()` exposes the **same API** regardless of which backend
you use — only the constructor differs:

```{r eval=FALSE}
backends <- list(
  ollama   = ollama_backend("llama3.2"),
  vllm     = vllm_backend("my-model", base_url = "http://localhost:8000/v1"),
  sglang   = sglang_backend("my-model", base_url = "http://localhost:30000/v1"),
  llamacpp = llamacpp_backend("my-model", base_url = "http://localhost:8080/v1")
)

purrr::imap(backends, ~ {
  classifier <- llm_classifier(.x)
  result <- classify(classifier, "Hello world!", choices = c("a", "b", "c"))
  paste0(.y, ": ", result$prediction)
})
```

## Inspecting a backend

Every backend is a list with the unified methods plus capability
metadata. This is primarily useful for debugging or advanced use:

```{r eval=FALSE}
backend <- ollama_backend("llama3.2")

backend$chat        # constrained chat completion
backend$score       # completion scoring (used by classify())
backend$tokenize    # context-dependent tokenization
backend$supports_bare_label_constraint  # FALSE for Ollama, TRUE elsewhere
```
