---
title: "Inference Backends"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{inference-backends}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Overview

`rollama` supports four inference backends. The original `ollama_classifier()`
uses the Ollama REST API directly. For all other backends, use `llm_classifier()`
with the appropriate backend class.

All non-Ollama backends communicate via the [OpenAI-compatible chat completions API](https://platform.openai.com/docs/api-reference/chat).

## Ollama

The default backend. Requires the Ollama runtime installed locally.

```{r eval=FALSE}
library(rollama)
classifier <- ollama_classifier("llama3.2")
```

## vLLM

High-throughput serving engine for LLMs. Supports guided decoding and logprobs out of the box.

**Start a local server:**

```bash
python -m vllm.entrypoints.openai.api_server \
    --model meta-llama/Llama-3.2-3B-Instruct \
    --host 0.0.0.0 --port 8000
```

**Connect:**

```{r eval=FALSE}
backend <- vllm_backend(
  model = "meta-llama/Llama-3.2-3B-Instruct",
  base_url = "http://localhost:8000/v1"
)
classifier <- llm_classifier(backend)
```

**Remote server:**

```{r eval=FALSE}
backend <- vllm_backend(
  model = "your-model",
  base_url = "https://your-vllm-server.com/v1",
  api_key = "your-api-key"
)
classifier <- llm_classifier(backend)
```

## SGLang

Fast serving system for large language models with efficient radix attention.

**Start a local server:**

```bash
python -m sglang.launch_server \
    --model-path meta-llama/Llama-3.2-3B-Instruct \
    --host 0.0.0.0 --port 30000
```

**Connect:**

```{r eval=FALSE}
backend <- sglang_backend(
  model = "meta-llama/Llama-3.2-3B-Instruct",
  base_url = "http://localhost:30000/v1"
)
classifier <- llm_classifier(backend)
```

## llama.cpp

Lightweight inference via `llama-server`. Ideal for CPU or mixed CPU/GPU environments.

**Start a local server:**

```bash
./llama-server -m model.gguf --host 0.0.0.0 --port 8080 -c 4096
```

**Connect:**

```{r eval=FALSE}
backend <- llamacpp_backend(
  model = "model",
  base_url = "http://localhost:8080/v1"
)
classifier <- llm_classifier(backend)
```

> **Note:** JSON schema constraints and logprobs require llama.cpp to be
> compiled with the appropriate flags (`LLAMA_JSON_SCHEMA` and
> `LLAMA_SUPPORT_LOGPROBS`).

## Backend Configuration

All backends share common configuration options:

| Parameter    | Default        | Description                                |
|--------------|----------------|--------------------------------------------|
| `model`      | *(required)*   | Model identifier                           |
| `base_url`   | Engine-specific | Base URL of the inference server           |
| `api_key`    | `"not-needed"` | API key for authentication                 |
| `timeout`    | `120`          | Request timeout in seconds                 |
| `max_tokens` | `256`          | Maximum tokens to generate                 |
| `extra_body` | `{}`           | Extra parameters merged into every request |

## Switching Backends

The `llm_classifier()` exposes the **same API** regardless of which backend you use:

```{r eval=FALSE}
backends <- list(
  vllm = vllm_backend("my-model", base_url = "http://localhost:8000/v1"),
  sglang = sglang_backend("my-model", base_url = "http://localhost:30000/v1"),
  llamacpp = llamacpp_backend("my-model", base_url = "http://localhost:8080/v1")
)

purrr::imap(backends, ~ {
  classifier <- llm_classifier(.x)
  result <- classify(classifier, "Hello world!", choices = c("a", "b", "c"))
  paste0(.y, ": ", result$prediction)
})
```
